---
title: "ggschemes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggschemes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggschemes)
library(ggplot2)
```

This vignette introduces the basics of (1) using `ggschemes`' built-in themes and scales with `ggplot2`-objects, and (2) building your own themes, color schemes, palettes & color scales using the tools `ggschemes` provides.

# Using pre-built themes and color scales

The themes and scales provided by `ggschemes` can be used mostly interchangeably with other themes and scales for `ggplot2`. Out of the box, `ggschemes` provides three color schemes, each coming with four themes: `theme_*_light()`, `theme_*_light_horizontal()`, `theme_*_dark()`, and `theme_*_dark_horizontal()`. They only have one argument, `...`, which is used to pass all arguments given directly to `theme()`. All color schemes also come with two scale functions: `scale_color_*()`, for color, and `scale_fill_*()`, for fill aesthetics. They have three arguments: `palette`, for the color palette (defaults to "light"), `discrete`, taking a logical value, for choosing discrete vs. continuous color scales (I think this is more convenient than separate functions), and `reverse`, in case you want to reverse the palette colors. Let's look at some examples!

## Building a simple plot

The themes can simply be added to a `ggplot()`-call as a layer, like any other theme:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg)) +
  geom_point() +
  theme_dracula_light()          # <-
```

To use the color scales, like with any color scale, you first add an aesthetic mapping for color, and then add the color scale as extra layer:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = factor(cyl))) + # <-
  geom_point() +
  theme_dracula_light() +
  scale_color_dracula()                               # <-
```

Let's say you switch to the dark theme; you may want to switch to the dark palette as well:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = factor(cyl))) + 
  geom_point() +
  theme_dracula_dark() +                              # <-
  scale_color_dracula(palette = "dark")               # <-
```


The color scale can also display continuous values. You may want to pick a palette that works well with continuous values (the function will automatically interpolate between color values). 

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = mpg)) +         # <- 
  geom_point() +
  theme_dracula_dark() +
  scale_color_dracula(palette = "rank", discrete = F) # <-  
```

You can also manually create your own scales by accessing individual colors from the color scheme. To check what colors are available, simply call the name of the color scheme:

```{r}
dracula()
```

One exception here is `gruvbox`, which has a light and a dark variant:

```{r}
gruvbox$light()
gruvbox$dark()
```

For example, you can create you own gradient color scale:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = mpg)) +
  geom_point() +
  theme_dracula_dark() +
  scale_color_gradient(low = dracula("pink"), high = dracula("cyan")) # <- 
```

# Creating your own color schemes, palettes & themes

`ggschemes` also provides a number of convenience functions to make creating your own color schemes, palettes & themes much easier. This section will introduce these ideas. First, create a named vector holding the color names and hexcodes for your palette:

```{r}
my_colors <- c(
  `black` = "#2e2e2e",
  `white` = "#d6d6d6", 
  `yellow` = "#e5b567",
  `green` = "#b4d273",
  `orange` = "#e87d3e",
  `purple` = "#9e86c8",
  `pink` = "#b05279",
  `blue` = "#6c99bb",
  `grey` = "#797979"
)
```

Next, you can use the `make_colorscheme()` function factory to create a function which accesses the colors:

```{r}
my_color <- make_colorscheme(my_colors)
```

You can now access the colors from your palette by name:

```{r}
my_color("purple")
my_color("purple", "pink")
my_color()
```

Next, you can assemble some palettes from your colors & put them into a list:

```{r}
my_palettes <- list(
  `light` = my_color("yellow", "green", "orange", "purple", "pink", "blue"),
  `dark` = my_color("yellow", "green", "orange", "purple", "pink", "blue"),
  `cold` = my_color("white", "blue"),
  `warm` = my_color("yellow", "pink"),
  `rank` = my_color("pink", "green")
)
```

You can now create a function to access these palettes:

```{r}
my_pal <- make_palette(my_palettes)
```

...which you can use to construct color & fill scales:

```{r}
my_color_scale <- make_color_scale(
  .palette_fun = my_pal,
  .name_pref = "my_"
)

my_fill_scale <- make_fill_scale(
  .palette_fun = my_pal,
  .name_pref = "my_"
)
```

We can now already use the color & fill scales with `ggplot()`:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  my_color_scale()  # <- 
```

Let's create a theme. `ggschemes` includes two factories: `make_dark_theme()` and `make_light_theme()`. We will stick with the light theme for now. The function has a `.template` argument to select a template to base the theme on. `ggschemes` comes with two built-in templates, `theme_template()` and `theme_template_horizontal()`, but you can use any theme as template or create your own using `ggplot2::theme()`. You will also need to specify a number of colors:

```{r}
my_light_theme <- make_light_theme(
  .template = theme_template(),
  .black = my_color("black"),
  .grey = my_color("grey"),
  .lgrey = my_color("grey"),  # sometimes a separate light grey is nice for contrast!
  .white = my_color("white")
)
```

The theme can now be used with a plot:

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  my_light_theme() +  # <- 
  my_color_scale()
```

If you dislike the templates, you can either create your own template, or overwrite parts of the theme with a separate call to `ggplot2::theme()` as extra layer (for example, removing axis lines & ticks):

```{r}
mtcars |> 
  ggplot(aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  my_light_theme() +
  my_color_scale() +
  theme(                          # <- 
    axis.line = element_blank(),  # <- 
    axis.ticks = element_blank()  # <- 
  )                               # <- 
```

All functions are written with flexibility and compatibility with other `ggplot2` functions in mind, to enable these kinds of things. If you want to extend `ggschemes` or use it to build your own themes & color palettes, I suggest you also look at the source code for the existing ones; the basic workflow is identical to what was presented here. 